Efficient pedigree recording for fast population genetics simulation

### 2017-12-30 forwards_paper.tex
Formatting tweak for msprime method/class names

\begin{abstract}
    In this paper we describe how to
    efficiently record the entire genetic history of a population
    during a forwards-time, individual-based population genetics simulation.
    This allows us to dramatically reduce the computational burden of tracking individual genomes
    by simulating only those loci that may affect reproduction (those having non-neutral variants),
    and recording history in the \emph{tree sequence} format introduced in the software package \msprime,
    on which neutral mutations can be quickly placed afterwards.
    We implement the method in two popular forwards-time simulation frameworks,
    and show that not only does this speed up large simulations by one or two orders of magnitude,
    but saves substantially more information from the simulation
    (all marginal genealogies) than simply recording the final sequences of
    simulated individuals.
    \jda{ this is still only 120 words and we should say 'more than what'}
    % This has the promise of making large-scale, whole-genome simulations with realistic geography and selection finally possible.
\end{abstract}

@@

Since the 1980's, coalescent theory has enabled computer simulation of the results of population genetics models
identical to that which would be produced by large, randomly mating populations over long periods of time
without actually requiring simulation of so many generations or meioses.
Coalescent theory thus had three transformative effects on population genetics:
first, giving researchers better conceptual tools to describe \emph{gene trees} and thus bringing within-population trees into better focus;
second, producing analytical methods to estimate parameters of interest from genetic data (e.g., $\theta = 4N_e \mu$);
and finally, providing a computationally feasible method to produce computer simulations of population genetics processes.
However, these powerful advances came with substantial caveats:
the backwards-in-time processes that are described by coalescent theory
are only \emph{Markovian}, and thus feasible to work with,
thanks to the important assumptions of
(a) random mating,
(b) neutrality,
and (c) small sample size relative to the population size.
The first two assumptions can be side-stepped to a limited extent \jda{cite?},
but it remains a challenge to map results of coalescent models
onto species that are distributed across continuous geographical space
and/or have large numbers of loci under various sorts of selection.
Furthermore, the relationship between the life history of a species --
fecundity and mortality schedules, density-dependent effects on fitness, and demographic fluctuations --
are all absorbed into a single compound parameter, the coalescence rate.
The third assumption is no longer safe, either --
for example, a recent study~\citep{martin2017human}
simulated 600,000 samples of human chromosome 20 to examine biases in GWAS.
Several studies have now shown that in samples of size approaching that of the population,
genealogical properties may be distorted relative to the coalescent expectation
\citep{wakeley2003gene,maruvka2011recovering,bhaskar2014distortion}.
These considerations, and increasing computational power, have led to a resurgence of
interest in large forwards-time, individual-based simulations.
For instance, \citet{harris2016genetic} used SLiM \citep{slim} to simulate ten thousand human exomes
to assess the impact of genetic load and Neanderthal introgression on human genetic diversity.
\cite{Sanjak2017-ko} used
fwdpp \citep{fwdpp} to simulate a series of models of quantitative traits under mutation-selection balance with
population sizes of $2 \times 10^4$ diploids in stable populations and populations growing up to $\sim 5
\times 10^5$ individuals, using the output to explore the relationship between the genotype/phenotype model and GWAS
outcomes.

Modern computing power easily allows simulations of birth, death and reproduction
in a population having even hundreds of millions of individuals.
However, if our interest lies in the resulting genetic patterns of variation
-- and often, the point of such simulations is to compare to real data --
then such simulations must record data for each individual genome.
As samples of most species' genomes harbor tens or hundreds of millions of variant sites,
naively carrying full genotypes for even modest numbers of individuals through a simulation
becomes quickly prohibitive.
To make matters worse,
a population of size $N$ must be simulated across many multiples of $N$ generations
to produce stable genetic patterns \jda {citation?}.
Because of this computational burden, even the fastest simulation frameworks such as SLiM,
\jda{We haven't motivatd that neutral muts are the burden yet}
SLiM2~\citep{haller2017flexible} and fwdpp~\citep{fwdpp}
can ``only'' simulate tens of megabases of sequence in tens of thousands of individuals
for tens of thousands of generations, and there are still near-order-of-magnitude performance differences between these
implementations~\citep{haller2017flexible}.
In practice, current state-of-the-art simulation software may take on the order of
weeks to simulate models of large genomic regions without selection~\citep{fwdpp,Hernandez2015-wf},
and existing simulation engines differ in how efficiently they
calculate fitnesses in models with selection~\citep{fwdpp}.
These population and region sizes are still substantially short of whole genomes
(hundreds to thousands of megabases)
for many biological population sizes of interest.

However, it is thought that most genetic variation is selectively neutral (or nearly so).
By definition, the neutral alleles carried by individuals in a population
do not affect the population process.
For this reason, if one records the entire genealogical history of a population over the course of a simulation,
simply laying down neutral mutations on top of that history afterwards
is equivalent to having generated them during the simulation.
Precisely, we would need to know the genealogical tree relating all sampled individuals
at each position along the genome.
This \emph{population pedigree} is 
the complete history of parent-offspring relationships of an entire population
going back to a remote time including the genetic outcomes of each ancestral
meiosis.
The embellished pedigree contains all the information necessary
to construct the genealogical tree that relates each individual to every other
at each position on the genome,
i.e., the \emph{tree sequence} implemented in the coalescent simulator \msprime{}~\citep{kelleher2016efficient}.
\jda{Currently (even before my edit) we no longer mention the ARG equivalence *at all* in the intro which
	seems wrong, but we might've hashed this over already? (and it's
	mentioned later so maybe no matter)}
Combined with ancestral genotypes and the origins of new mutations,
it also completely specifies the genomic sequence of any individual in the population at any time.
This is much much more than typically recorded in a forward simulation, however,
so further savings can be had by discarding all information irrelevant to the genetic history
of the extant population.
\jda{The space complexity result seemed like a result in the intro; and maybe didn't include some detail
	about assumed number of generations}

In this paper, we describe storage methods for genealogies
(and hence, genome sequence) and an algorithm for simplifying these,
both implemented as improvements to the algorithmic tools and data structures in \msprime{}.
We also describe a general method of using these tools to efficiently record,
and later process, the population pedgiree in forwards-time simulation.
While providing substantial savings in time and space, our methods provide much
more information than simply simulating the genomes -- the tree sequence format
encodes all marginal genealogies of individuals living at the end of the
simulation.
Although we were motivated by a need for more efficient genomic simulations,
these tools may prove more widely useful.

>>>>

\begin{abstract}
    In this paper we describe how to
    efficiently record the entire genetic history of a population
    during a forwards-time, individual-based population genetics simulation.
    This allows us to dramatically reduce the computational burden of tracking individual genomes
    by simulating only those loci that may affect reproduction (those having non-neutral variants),
    and recording history in the \emph{tree sequence} format introduced in the software package \msprime,
    on which neutral mutations can be quickly placed afterwards.
    We implement the method in two popular forwards-time simulation frameworks,
    and show that not only does this speed up large simulations by one or two orders of magnitude,
    but saves substantially more information from the simulation
    (all marginal genealogies).
    % This has the promise of making large-scale, whole-genome simulations with realistic geography and selection finally possible.
\end{abstract}

@@

Since the 1980's, coalescent theory has enabled computer simulation of the results of population genetics models
identical to that which would be produced by large, randomly mating populations over long periods of time
without actually requiring simulation of so many generations or meioses.
Coalescent theory thus had three transformative effects on population genetics:
first, giving researchers better conceptual tools to describe \emph{gene trees} and thus bringing within-population trees into better focus;
second, producing analytical methods to estimate parameters of interest from genetic data (e.g., $\theta = 4N_e \mu$);
and finally, providing a computationally feasible method to produce computer simulations of population genetics processes.
However, these powerful advances came with substantial caveats:
the backwards-in-time processes that are described by coalescent theory
are only \emph{Markovian}, and thus feasible to work with,
thanks to the important assumptions of
(a) random mating,
(b) neutrality,
and (c) small sample size relative to the population size.
The first two assumptions can be side-stepped to a limited extent,
but it remains a challenge to map results of coalescent models
onto species that are distributed across continuous geographical space
and/or have large numbers of loci under various sorts of selection.
Furthermore, the relationship between the life history of a species --
fecundity and mortality schedules, density-dependent effects on fitness, and demographic fluctuations --
are all absorbed into a single compound parameter, the coalescence rate.
The third assumption is no longer safe, either --
for example, a recent study~\citep{martin2017human}
simulated 600,000 samples of human chromosome 20 to examine biases in GWAS.
Several studies have now shown that in samples of size approaching that of the population,
genealogical properties may be distorted relative to the coalescent expectation
\citep{wakeley2003gene,maruvka2011recovering,bhaskar2014distortion}.
These considerations, and increasing computational power, have led to a resurgence of
interest in large forwards-time, individual-based simulations.
For instance, \citet{harris2016genetic} used SLiM \citep{slim} to simulate ten thousand human exomes
to assess the impact of genetic load and Neanderthal introgression on human genetic diversity.
\cite{Sanjak2017-ko} used
fwdpp \citep{fwdpp} to simulate a series of models of quantitative traits under mutation-selection balance with
population sizes of $2 \times 10^4$ diploids in stable populations and populations growing up to $\sim 5
\times 10^5$ individuals, using the output to explore the relationship between the genotype/phenotype model and GWAS
outcomes.

Modern computing power easily allows simulations of birth, death and reproduction
in a population having even hundreds of millions of individuals.
However, if our interest lies in the resulting genetic patterns of variation
-- and often, the point of such simulations is to compare to real data --
then such simulations must somehow produce at the end data for each individual on a genomic scale.
As samples of most species' genomes harbor tens or hundreds of millions of variant sites,
naively carrying full genotypes for even modest numbers of individuals through a simulation
becomes quickly prohibitive.
To make matters worse,
a population of size $N$ must be simulated across many multiples of $N$ generations
to produce stable genetic patterns.
Because of this computational burden of neutral mutations, even the fastest simulation frameworks such as SLiM,
SLiM2~\citep{haller2017flexible} and fwdpp~\citep{fwdpp}
can ``only'' simulate tens of megabases of sequence in tens of thousands of individuals
for tens of thousands of generations, and there are still near-order-of-magnitude performance differences between these
implementations~\citep{haller2017flexible}.
In practice, current state-of-the-art simulation software may take on the order of
weeks to simulate models of large genomic regions without selection~\citep{fwdpp,Hernandez2015-wf},
and existing simulation engines differ in how efficiently they
calculate fitnesses in models with selection~\citep{fwdpp}.
These population and region sizes are still substantially short of whole genomes
(hundreds to thousands of megabases)
for many biological population sizes of interest.

However, it is thought that most genetic variation is selectively neutral (or nearly so).
By definition, the neutral alleles carried by individuals in a population
do not affect the population process.
For this reason, if one records the entire genealogical history of a population over the course of a simulation,
simply laying down neutral mutations on top of that history afterwards
is equivalent to having generated them during the simulation.
Precisely, we would need to know the genealogical tree relating all sampled individuals
at each position along the genome.
In this paper, we describe how algorithmic tools and data structures developed for the
coalescent simulator \msprime{}
can be used to efficiently record, and later process, this history.
In so doing we record the \emph{population pedigree} --
the complete history of parent-offspring relationships of an entire population
going back to a remote time including the genetic outcomes of each ancestral
meiosis.
This embellished pedigree contains all the information necessary
to construct the genealogical tree that relates each individual to each other
at each position on the genome, i.e., the \emph{tree sequence}~\citep{kelleher2016efficient}.
Combined with ancestral genotypes and the origins of new mutations,
it also completely specifies the genomic sequence of any individual in the population at any time.
This is much more than we need to know, however,
so we discard all information irrelevant to the genetic history
of the extant population. We show that we need only $O(N \log N)$
space to store the resulting tree sequence for a single crossover per
generation under the Wright-Fisher model.

In this paper we discuss storage methods for genealogies (and hence, genome sequence),
an algorithm for simplifying these,
and their use in forwards-time simulation.
Although we were motivated by a need for more efficient genomic simulations,
these tools may prove more widely useful.
